```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5,
  dpi        = 110,
  cache      = TRUE,
  cache.lazy = FALSE
)

GLOBAL_SEED <- 42
set.seed(GLOBAL_SEED)

options(
  digits     = 4,
  scipen     = 999,
  dplyr.summarise.inform = FALSE
)
```

> **Submission.** Knit your Rmd file into HTML and submit it via Brightspace. Include a link to your git repository (GitHub or git.cs.dal.ca) as a comment on your submission.


Name: Malk Daboor
email: malk.daboor@dal.ca
B00: B00953701
GitHub link: https://github.com/serendipity77705/csci4148-labs.git
GitHub Rmd link: https://github.com/serendipity77705/csci4148-labs/blob/main/lab1_medical_databases.rmd
---

## 0 Setup

```{r packages}
library(tidyverse)    # dplyr, ggplot2, tidyr, purrr, ...
library(tidymodels)   # rsample, recipes, parsnip, workflows, tune, yardstick, ...
library(ranger)       # random forest engine
library(xgboost)      # gradient-boosted trees engine
library(glmnet)       # penalised logistic regression engine
library(here)
library(fs)

theme_set(theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank()))
```

---

## 1 Introduction

[Sepsis](https://www.who.int/news-room/fact-sheets/detail/sepsis) is life-threatening organ dysfunction caused by a dysregulated immune response to infection.
It accounts for roughly 20 % of global deaths, and each hour of delayed treatment increases mortality by 4–8 %.

The [PhysioNet/CinC Challenge 2019](https://physionet.org/content/challenge-2019/1.0.0/) dataset contains hourly ICU records for ~40 000 patients from two US teaching hospitals.
Each record (row) has 8 vital signs, 26 laboratory measurements, 6 demographic fields, and a binary `SepsisLabel` (set to 1 up to 6 hours *before* clinical onset so that early-warning models can be trained).
We simplify the temporal problem to a static one by taking one random hourly record per patient, giving a standard tabular classification task with realistic clinical challenges: class imbalance, high missingness, and mixed data types.

**Q1. (2 marks)** Look up the [PhysioNet variable descriptions](https://physionet.org/content/challenge-2019/1.0.0/) and include an explanation of the dataset into this notebook (one row per variable: name, description, units).  
the dataset includes vital signs like heart rate, temperature, respiration rate, etc.; labratory values like lactic acid amount; demographics like age and gender; and the final outcome.

HR	Heart rate (beats per minute)
O2Sat	Pulse oximetry (%)
Temp	Temperature (Deg C)
SBP	Systolic BP (mm Hg)
MAP	Mean arterial pressure (mm Hg)
DBP	Diastolic BP (mm Hg)
Resp	Respiration rate (breaths per minute)
EtCO2	End tidal carbon dioxide (mm Hg)

BaseExcess	Measure of excess bicarbonate (mmol/L)
HCO3	Bicarbonate (mmol/L)
FiO2	Fraction of inspired oxygen (%)
pH	N/A
PaCO2	Partial pressure of carbon dioxide from arterial blood (mm Hg)
SaO2	Oxygen saturation from arterial blood (%)
AST	Aspartate transaminase (IU/L)
BUN	Blood urea nitrogen (mg/dL)
Alkalinephos	Alkaline phosphatase (IU/L)
Calcium	(mg/dL)
Chloride	(mmol/L)
Creatinine	(mg/dL)
Bilirubin_direct	Bilirubin direct (mg/dL)
Glucose	Serum glucose (mg/dL)
Lactate	Lactic acid (mg/dL)
Magnesium	(mmol/dL)
Phosphate	(mg/dL)
Potassium	(mmol/L)
Bilirubin_total	Total bilirubin (mg/dL)
TroponinI	Troponin I (ng/mL)
Hct	Hematocrit (%)
Hgb	Hemoglobin (g/dL)
PTT	partial thromboplastin time (seconds)
WBC	Leukocyte count (count*10^3/µL)
Fibrinogen	(mg/dL)
Platelets	(count*10^3/µL)

Age	Years (100 for patients 90 or above)
Gender	Female (0) or Male (1)
Unit1	Administrative identifier for ICU unit (MICU)
Unit2	Administrative identifier for ICU unit (SICU)
HospAdmTime	Hours between hospital admit and ICU admit
ICULOS	ICU length-of-stay (hours since ICU admit)

SepsisLabel	For sepsis patients, SepsisLabel is 1 if t≥tsepsis−6 and 0 if t<tsepsis−6. For non-sepsis patients, SepsisLabel is 0.

**Q2. (2 marks)** Then choose **three variables** you hypothesise might be strong predictors of sepsis and write one sentence each explaining your reasoning.  There is no right or wrong answer to this, but you should think about your and problem before diving into the EDA and prediction!
*(Hint: look up how sepsis is clinically diagnosed.)*  

The three varuables I will be using are lactate level, heart rate and blood urea levels.
- lactate: lactic acid is secreted when your body isn't receiving enough oxygen due to heart failure, which sepsis can cause, or severe infection, which is a sign of possible sepsis. High lactic acid values may indicate sepsis.
- heart rate: is a visible sign that something is worng. Often the heart beats irregularly fast when the body has sepsis
- respiration rate: sign that something is wrong if patient has shallow breaths and body is trying to get oxygen into it.
---

## 2 Data Loading and Splitting

To avoid inadvertent data leakage from your test set, we are going to perform our test-train split before even our EDA and cleaning related analyses. This is not always done but is a good practice if you aren't using an external/independent test dataset.

### 2.1 Load the data

I've already done the random sampling of the dataset to remove the time-series elements but the original dataset can be acquired directly from [PhysioNet/CinC Challenge 2019](https://physionet.org/content/challenge-2019/1.0.0/) 
```{r data-paths}
data_dir  <- here("data")
cache_dir <- fs::path(data_dir, "cache")
fs::dir_create(cache_dir)

data_url <- "http://maguire-lab.github.io/health_data_science_research_2026/static_files/practicals/lab1_data/physionet_subsample_modified_1.0.0.rds"
rds_path <- fs::path(cache_dir, "physionet_subsample_1.0.0.rds")
```

Rmd lets you cache expensive or slow operations (like downloading big files or complicated processing) so they only re-run if the code and variables have changed.  This can be very useful although can sometimes cause some inadvertent issues that reduce reproducibility!  You should also do a completely clean re-run of everything once your analyses are finalised if possible.

```{r load-data, cache = TRUE}
if (!file.exists(rds_path)) {
  download.file(data_url, rds_path, mode = "wb", quiet = TRUE)
}
sepsis <- readRDS(rds_path)
```

### 2.2 Train / test split

We split the data **before** any exploration so that analysis choices cannot be influenced by the test set - this is a key discipline in applied machine learning.  Stratified splitting preserves the sepsis rate in both halves.

```{r split}
# will make later things easier to do!s
sepsis <- sepsis |>
  mutate(SepsisLabel = factor(SepsisLabel,
                              levels = c(1, 0),
                              labels = c("Sepsis", "No Sepsis")))

# splitting
split <- initial_split(sepsis, prop = 0.75, strata = SepsisLabel) # 75% of data is training, 25% is testing
train <- training(split)
test  <- testing(split)

cat("Training rows:", nrow(train), "\n")
cat("Test rows:    ", nrow(test),  "\n")
```

**Q3. (1 mark)** What percentage of our dataset is being assigned to test?
25% of data is assigned to test, while 75% is training
--- 

## 3 Exploratory Data Analysis

Again, do your EDA on `train` only - we do not look at `test` until Section 6.

### 3.1 Class balance


**Q4. (2 marks).** Uncomment and complete the code below to check the class balance in the **training** set.
`count()` tallies rows per group; add a `prop` column to show each class as a fraction of the total.

```{r q4-class-balance}
train |>
  count(SepsisLabel) |>
  mutate(prop = n / sum(n))
```


**Q5. (2 mark)** Using your solution to Q4, complete this plotting code to create a barplot of the relative sepsis label balance.  Change the colour of the sepsis bar to green.

```{r class-balance-plot, fig.height = 2.8}
#code from your answer to Q4,
train |>
  count(SepsisLabel) |>
  mutate(prop = n / sum(n)) |>

  ggplot(aes(SepsisLabel, prop, fill = SepsisLabel)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.01)), vjust = -0.4) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("No Sepsis" = "grey60", 
                               "Sepsis"    = "green")) +
  labs(x = NULL, y = "Percentage of Trainining Data", title = "Sepsis Class Balance") +
  theme(legend.position = "none")
```

**Q6. (3 marks)**  Given the class imbalance, which of the following 2 metrics might be the most misleading and why? accuracy, precision, recall, F1-score, ROC-AUC, PR-AUC?
The 2 metrics that would be misleading are accuracy, and ROC-AUC.

Just for my knowledge: 
-- accuracy: tells us number of correct predictions over number of total predictions. With imbalanced datasets, we might have high accuracy because class A is being predicted correctly all the time while class B is being predicted correctly none of the time. However, the inbalance for A will still lead to high accuracy
-- precision: measures how many of the positive predictions are correct. Good for when we don't want false positives ie medicine. 
-- recall: opposite of precision, where we care about false negatives more than false positives. 
-- F1-score: used when we want balance between precision and recall
-- ROC-AUC: binary classification model, is for sensitivity (true positives) vs. specificity (false positives). it is good for balanced datasets 
-- PR-AUC: binary classification model, is for precision (positive predictions) vs. recall (true positives). Is good for unbalanced datasets


### 3.2 Missingness

```{r missingness-bar, fig.height = 5}
train |>
  summarise(across(everything(), ~ mean(is.na(.x)))) |>
  pivot_longer(everything(),
               names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  ggplot(aes(pct_missing, fct_reorder(variable, pct_missing))) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Missing data by variable (training set)",
       x = "Percent missing", y = NULL)
```

**Q7. (1 mark)**
Modify the above to plot only missingness for Sepsis positive data
```{r missingness-sepsis-bar, fig.height = 5}
# add your code
train |>
  filter(SepsisLabel == "Sepsis") |>
  summarise(across(everything(), ~ mean(is.na(.x)))) |>
  pivot_longer(everything(),
               names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  ggplot(aes(pct_missing, fct_reorder(variable, pct_missing))) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(title = "Missing data by variable (training set) for Sepsis positive data",
       x = "Percent missing", y = NULL)

```

**Q8. (3 marks)** Which three variables have the highest missingness? In an ICU context, why might a lab value be missing - is this likely to be random, or informative? Does this mean you should drop or impute these values?
The 3 variables are:
- troponinI: Troponin I (ng/mL) -- protein that is found in heart muscles and helps with heart contractions
- fibrinogen: (mg/dL) -- protein that circulates in the body and clots blood around injuries to stop bleeding
- bilirubin_direct: Total bilirubin (mg/dL) -- pigment that is created when old red blood cells are broken down

the reason these might have a high missingness is because there might be faster and easier variables to collect, and this would be quite important in severe cases where there might not be enough time to consistently test for these values. They also might not be as impactful as other variables such as heart and respiration rate. We have no/almost no data on these three variables so it would not be very useful to have them; it would be better to drop them and focus on morei mpactful variables.


### 3.3 Vital-sign distributions by outcome

```{r vitals-density, fig.height = 5}
vitals <- c("HR", "O2Sat", "Temp", "SBP", "MAP", "DBP", "Resp")

train |>
  select(all_of(vitals), SepsisLabel) |>
  pivot_longer(-SepsisLabel) |>
  filter(!is.na(value)) |>
  ggplot(aes(value, fill = SepsisLabel)) +
  geom_density(alpha = 0.55, colour = NA) +
  facet_wrap(~ name, scales = "free", ncol = 4) +
  scale_fill_manual(values = c("No Sepsis" = "grey60",
                               "Sepsis"    = "firebrick")) +
  labs(x = NULL, y = "Density", fill = NULL,
       title = "Vital-sign distributions by sepsis status")
```

**Q9 (2 marks).** Modify the code above to plot the distribution of **three other lab variables** of your choice. 

```{r q9-lab-density}
vitals <- c("HR", "O2Sat", "Temp", "SBP", "MAP", "DBP", "Resp", "pH", "Glucose", "BaseExcess") # added pH, glucose and excess bicarbonate

train |>
  select(all_of(vitals), SepsisLabel) |>
  pivot_longer(-SepsisLabel) |>
  filter(!is.na(value)) |>
  ggplot(aes(value, fill = SepsisLabel)) +
  geom_density(alpha = 0.55, colour = NA) +
  facet_wrap(~ name, scales = "free", ncol = 4) +
  scale_fill_manual(values = c("No Sepsis" = "grey60",
                               "Sepsis"    = "firebrick")) +
  labs(x = NULL, y = "Density", fill = NULL,
       title = "Vital-sign distributions by sepsis status")
```

Correlation plot of all variables by similarity of their correlation in Sepsis and Non-Sepsis

```{r}
cor_mat <- function(df) {
  df |>
    select(where(is.numeric)) |>           
    cor(use = "pairwise.complete.obs")     
}

m_ns  <- cor_mat(train |> filter(SepsisLabel == "No Sepsis"))
m_sep <- cor_mat(train |> filter(SepsisLabel == "Sepsis"))
m_dif <- m_sep - m_ns

m_dif[is.na(m_dif)] <- 0
hc <- hclust(as.dist(1 - abs(m_dif)), method = "average")
var_order <- hc$labels[hc$order]

m_dif |>
  as.data.frame() |>
  tibble::rownames_to_column("var1") |>
  tidyr::pivot_longer(-var1, names_to = "var2", values_to = "diff") |>
  mutate(var1 = factor(var1, levels = var_order),
         var2 = factor(var2, levels = var_order)) |>
  ggplot(aes(var1, var2, fill = diff)) +
  geom_tile() +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0, limits = c(-0.5, 0.5),
                       oob = scales::squish, name = "Δ r") +
  coord_fixed() +
  labs(x = NULL, y = NULL,
       title = "Change in pairwise correlations: Sepsis − No Sepsis") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid  = element_blank())
```

### 3.4 Dimensionality reduction

PCA compresses all numeric features into a small number of axes and lets us ask: *do septic and non-septic patients separate at all in the joint feature space?*
We build it inside a small `recipe` so the preprocessing (imputation, normalisation) is explicit and reproducible.

Recipes are a special type of function in `tidymodels` that differ from normal R functions in key ways (internal state).

A plain R function executes immediately and is stateless i.e., every call recomputes from scratch.
```
rscale_hr <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)

train$HR_z <- scale_hr(train$HR)
test$HR_z  <- scale_hr(test$HR)   # BUG: uses test set's own mean/sd
```

The test set is standardised using statistics computed *from the test set*, which leaks information and means train and test are now on different scales. To do it correctly you'd have to manually save `mean(trainHR)` and `sd(trainHR)` and `sd(trainHR)` and `sd(trainHR)` and reuse them and then do the same for every other transformation, for every column, for imputation, encoding, etc. It gets unmanageable fast.

A recipe on the otherhand is a two-phase object: first you declare the steps, then you `prep()` it on training data (which estimates and stores any required parameters means, sds, imputation values, factor levels, PCA loadings…), then you `bake()` it on any new data to apply the same learned transformations.

```{r pca-prep}
pca_prep <- recipe(SepsisLabel ~ ., data = train) |>
  step_rm(ICULOS, HospAdmTime) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_pca(all_numeric_predictors(), num_comp = 10, id = "pca") |>
  prep()

pca_scores <- bake(pca_prep, new_data = NULL)
```

```{r pca-scatter, fig.width = 5.5, fig.height = 4}
pca_scores |>
  ggplot(aes(PC01, PC02, colour = SepsisLabel)) +
  geom_point(alpha = 0.75, size = 0.7) +
  scale_colour_manual(values = c("No Sepsis" = "grey60",
                                 "Sepsis"    = "firebrick")) +
  labs(title = "PCA: PC1 vs PC2", colour = NULL) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 2)))
```

```{r pca-scree, fig.height = 3}
tidy(pca_prep, id = "pca", type = "variance") |>
  filter(terms == "percent variance") |>
  ggplot(aes(component, value)) +
  geom_col(fill = "steelblue") +
  labs(x = "Principal component", y = "% variance explained",
       title = "PCA scree plot")
```


The two classes overlap heavily in the PC1/PC2 projection — this is common in clinical data where no single linear combination cleanly separates cases.

**Q10 (2 marks).** What does this tell us about the potential decision boundary our predictive model would have to fit?  Which models easily support this type of boundary?
Our model will have. to be very specific because of the large number of overlaps. We will not be able to use something as simple as a linear regression, rather we will have to consider other decision boundaries such as random forests, neural networks and support vector machines.  
---

## 4 Feature Engineering

Let's create a few new features likely to be informative for sepsis:

**Q11. (1 mark)** Modify the following code to also add an additional feature called `shock_index` which is defined as heart rate divided by the systolic blood pressure. 

```{r feature-engineering}
add_features <- function(df) {
  df |>
    mutate(
      pulse_pressure  = SBP - DBP,
      bun_creat_ratio = BUN / Creatinine,  # > 20 suggests pre-renal AKI
      age_group = cut(Age,
                      breaks = c(-1, 40, 65, 80, 120),
                      labels = c("<=40", "41-65", "66-80", "80+")),
      shock_index = HR/SBP
    )
}
train <- add_features(train)
test  <- add_features(test)
```

## 5 Modelling

Now lets try to fit some predictive models to the training set

### 5.1 Cross-validation folds

Five-fold stratified CV on the training set, used to select hyperparameters (see below) by repeatedly training the model with different hyperparameter values on 4/5ths of the data and evaluating on each of the held-out 1/5ths.s   

```{r cv-folds}
folds <- vfold_cv(train, v = 5, strata = SepsisLabel)
```

### 5.2 Preprocessing recipes

Not all models can handle the same type of data and some models (like linear models) benefit from normalising values to similar ranges. Most models will ignore missing values but we are going to add a missingness-indicator columns so the model can learn from *which* values are absent.

We are going to use `tidymodel` recipes again so we can apply the training normalisation etc to the test set to prevent leakage.

```{r recipes}
base_recipe <- recipe(SepsisLabel ~ ., data = train) |>
  step_rm(ICULOS, HospAdmTime) |>
  step_indicate_na(all_numeric_predictors(), prefix = "miss") |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_novel(all_nominal_predictors()) |>
  step_dummy(all_nominal_predictors(), one_hot = FALSE) |>
  step_zv(all_predictors())

glm_recipe <- base_recipe |>
  step_normalize(all_numeric_predictors())
```

### 5.3 Model specifications

Now let's specify our 3 models:
- Logistic regression: a linear baseline. Outputs P(Sepsis∣X) via a logistic link function (i.e., squeezing our linear model through a sigmoid function that constrains between 0 and 1). By default `glmnet` uses a regularised model.  This means we are penalising the model complexity (the size of the weights/coefficients are added to the cost function).  Specifically, we are using an cross-validation optimied combination of L1 and L2 penalties (absolute and squared weight/coefficient sizes).  This is useful as many of our ~40 lab variables are correlated or weakly informative.

- Random forest: an ensemble of decision trees, each fit on a bootstrap sample with a random subset of features at each split. Captures non-linearities and interactions automatically, and handles mixed-scale features without preprocessing. Robust to outliers and (via `ranger`) to NAs.

- XGBoost: gradient-boosted trees. Builds trees sequentially, each one correcting the residual errors of the previous ensemble. Typically the strongest tabular performer, but the most sensitive to model (hyper)parameters and easy to overfit without tuning.

Each of these has several hyperparameters (parts of the model we don't directly learn from the data during fitting; instead we choose them before fitting and select the best values by cross-validation). Marking a hyperparameter with `tune()` is a placeholder that says "leave this blank for now — we'll fill it in during the tuning step." The values will later be chosen by tune_grid() based on cross-validated performance on the training folds.

**Q12 4 marks**: Using R's documentation (e.g., `?logistic_reg`) look up and briefly explain each of the hyperparamters we are tuning in cross-validation for each of the 3 models:s

```{r model-spec}
# hyperparameters: penalty scaling factor and proportion L1 vs L2s
#penalty: total amount of regularization in model; refers to weights and penalties applied to a model's complexity to stop from overfitting.
#mixture: mixture amounts of different types of regularization
glm_spec <- logistic_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("classification")

# trees is set, not hyperparameter
# min_n: number of data points in node before it splits into other nodes
rf_spec <- rand_forest(trees = 500, min_n = tune()) |>
  set_engine("ranger", importance = "permutation", num.threads = 1) |>
  set_mode("classification")

#tree_depth: maximum depth of tree --> maximum number of splits
#learn_rate: iteration-to-iteration algorithm adaption rate
#loss_reduction: reduction in loss function required to split further
#sample_size: amount of data used when fitting
# min_n: number of data points in node before it splits into other nodes
xgb_spec <- boost_tree(
  trees = 500, tree_depth = tune(), learn_rate = tune(),
  loss_reduction = tune(), sample_size = tune(),
   min_n = tune()
) |>
  set_engine("xgboost", nthread = 1) |>
  set_mode("classification")
```


### 5.4 Tune all three models

Now let's specify our workflow to use cross-validation to tune all models and compare across them.

```{r workflow-set}
wf_set <- workflow_set(
  preproc = list(glm = glm_recipe, rf = base_recipe, xgb = base_recipe),
  models  = list(glm = glm_spec,   rf = rf_spec,      xgb = xgb_spec),
  cross   = FALSE
)
```
We are going to calculate a set of different performance metrics across CV folds.

- `pr_auc`: Area under the Precision-Recall curve. Sweeps the classification threshold from strict to lenient and plots precision against recall at each. Single-number summary of how well the model ranks septic patients above non-septic ones, without being inflated by the large true-negative pool. Baseline ≈ class prevalence (~0.02 here), so any value meaningfully above that is real signal. 

- `brier_class`: Mean squared error between predicted probabilities and the binary sepsis label 0/1 outcome. Measures **calibration + discrimination together** a model that says "80% sepsis risk" should be right 80% of the time. Lower is better; 0 is perfect, 0.25 is random guessing at 50/50. Unlike AUC metrics, Brier penalises *over-confident* probabilities

- `accuracy`: (TP+TN)/N. Due to imbalance this is included for familiarity and catching bugs! Reminder: Predicting "No Sepsis" for everyone scores ~98% here. 

- `sensitivity` (= recall, = true positive rate) TP/(TP+FN). Of patients who actually had sepsis, what fraction did we flag? The clinically critical metric for the positive class: missed sepsis (low sensitivity) means delayed antibiotics, which directly increases mortality.

- `specificity`: TN/(TN+FP). Of patients who did *not* have sepsis, what fraction did we correctly leave unflagged? The counterpart to sensitivity — high specificity means low false-alarm rate, which matters for **alarm fatigue** in the ICU. Sensitivity and specificity trade off as you move the threshold, so they're reported together.

Now we can run it, this will take a few minutes so we can use cache-ing again to avoid frustration!

```{r tune, cache = TRUE}
cls_metrics <- metric_set(pr_auc, brier_class, accuracy, sensitivity, specificity)

tuned <- wf_set |>
  workflow_map(
    fn        = "tune_grid",
    seed      = GLOBAL_SEED ,
    resamples = folds,
    grid      = 10,
    metrics   = cls_metrics,
    control   = control_grid(save_pred = TRUE, verbose = FALSE)
  )
```

### 5.5 Compare cross-validation results

```{r cv-comparison, fig.height = 3}
autoplot(tuned, metric = "pr_auc", select_best = TRUE) +
  labs(title = "Best CV PR-AUC per model family")
```

**Q13 (2 marks).** Which model family achieves the highest CV PR-AUC?
Is the gap between the best and worst model large relative to the standard errors?
What does that suggest about model selection here?
the model with the highest CV PR-AUC is logistic regression. random forest and boost trees seem to be at around the same CV PR-AUC values. the gap itself is not too big, at a difference of about 0.1, but the standard error seems like it would widen the gap.
It would be best to choose logistic regression as it performs the best out of all three models. it has the PR-AUC value and the smallest error bar, meaning that it is more consistent than the other two.

```{r}
rank_results(tuned, rank_metric = "pr_auc", select_best = TRUE) |>
  filter(.metric %in% c("pr_auc", "brier_class", "accuracy", "sensitivity", "specificity")) |>
  select(wflow_id, .metric, mean, std_err, rank)
```

**Q14 (3 marks).** By summarising and comparing the metrics output by this code, describe how each model performed relative to one another.
All three models ae quite similar to each other in terms of accuracy, brier class, PR-AUC and specificity. However, the main thing that they differ in is sensitivity, with gim_glm being 0.43, xgb_xgb at 0.5 and rf_rf at 0.34. All of these models are good at correctly identifying positive cases, but aren't as good at correctly identifying negative cases. In this specific case, xgb_xgb is a bit better than the other models, but it only about as good as a coin flip, with the other models being even less accurate than a random guess. 

---

## 6 Test-Set Evaluation

In the real-world we'd try to optimise these models further and/or try some different architectures but lets just take our best model and continue to test evaluation.

### 6.1 Finalise the best model

```{r finalise}
best_id <- rank_results(tuned, rank_metric = "pr_auc",
                        select_best = TRUE) |>
  filter(.metric == "pr_auc") |>
  slice_head(n = 1) |>
  pull(wflow_id)

final_wf <- extract_workflow(tuned, id = best_id) |>
  finalize_workflow(
    extract_workflow_set_result(tuned, id = best_id) |>
      select_best(metric = "pr_auc")
  )

split <- make_splits(train, test)
final_fit <- last_fit(final_wf, split = split, metrics = cls_metrics)
collect_metrics(final_fit)
```


```{r perf-curves, fig.width = 5.5, fig.height = 4}
preds <- collect_predictions(final_fit)

preds |>
  pr_curve(truth = SepsisLabel, .pred_Sepsis,
           event_level = "first") |>
  autoplot() + labs(title = "Precision–Recall curve (test set)")
```

### 6.3 Confusion matrix

**Q15 (1 marks).** How many times were true sepsis cases classified as non-sepsis in the test set? Is this more or less than the correct sepsis predictions?
18 patients were classified as non-sepsis when they had sepsis. The correct sepsis classifications are 11. This means that the sepsis cases classified as non-sepsis are more than correctly classified sepsis cases. The model misses more sepsis cases than it identifies. 

```{r conf-matrix}
preds |>
  conf_mat(truth = SepsisLabel, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(title = "Confusion matrix (threshold = 0.5)")
```

---

## 7 Misclassification Analysis

### 7.1 Subgroup performance

Now let's investigate whether the model work equally well/badly across patient groups?

```{r subgroup-auc, fig.height = 3.5}
errors <- augment(final_fit)

subgroup_auc <- function(df, group_var) {
  df |>
    group_by({{ group_var }}) |>
    filter(n_distinct(SepsisLabel) == 2) |>
    roc_auc(truth = SepsisLabel, .pred_Sepsis,
            event_level = "second") |>
    select({{ group_var }}, .estimate)
}

overall_auc <- collect_metrics(final_fit) |>
  filter(.metric == "pr_auc") |>
  pull(.estimate)

bind_rows(
  errors |> subgroup_auc(Gender)    |> rename(subgroup = 1) |>
    mutate(subgroup = as.character(subgroup), group = "Sex"),
  errors |> subgroup_auc(age_group) |> rename(subgroup = 1) |>
    mutate(subgroup = as.character(subgroup), group = "Age group")
) |>
  ggplot(aes(reorder(subgroup, .estimate), .estimate, fill = group)) +
  geom_col() +
  geom_hline(yintercept = overall_auc, linetype = 2, colour = "grey40") +
  coord_flip() +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  labs(x = NULL, y = "PR-AUC",
       title = "Subgroup PR-AUC",
       subtitle = "Dashed line = overall test AUC") +
  theme(legend.position = "none")
```

**Q15 (2 marks).** Is there a subgroup where the model performs noticeably worse than the overall PR-AUC?
Suggest one data-driven reason and one clinical reason that could explain the gap.
all the subgroups perform much worse than the overall PR-AUC. The absolute worse group in terms if performance though, is the 80+ age group, where the PR-AUC is 0.0 or just slightly more than that. One data-driven reason this could be the case is that there might just not be many data points collected about people that are 80+, so the PR-AUC would be bad just because of a lack of information. One clinical reason might be that for the elderly, they often have other medical conditions or complications that might have similar symptoms to sepsis, so they will not be classified as having sepsis but may be flagged for other conditions. Older patients also might not have the general symptoms of sepsis that a model is trained on, where they may not have a high fever or low blood pressure. Symptoms will present differently in the elderly comapred to younger populations. 

### 7.2 Variable importance

```{r variable-importance, fig.height = 10}
parsnip_fit <- extract_fit_parsnip(final_fit)

importance_tbl <- if (grepl("^rf", best_id)) {
  imp <- parsnip_fit$fit$variable.importance
  tibble(variable = names(imp), importance = unname(imp))
} else if (grepl("^xgb", best_id)) {
  xgboost::xgb.importance(model = parsnip_fit$fit) |>
    as_tibble() |>
    transmute(variable = Feature, importance = Gain)
} else {
  broom::tidy(parsnip_fit) |>
    filter(term != "(Intercept)") |>
    transmute(variable = term, importance = abs(estimate))
}

importance_tbl |>
  slice_max(importance, n = 20) |>
  ggplot(aes(importance, fct_reorder(variable, importance))) +
  geom_col(fill = "steelblue") +
  labs(title = paste("Top variable importances: ", best_id),
       x = "Importance", y = NULL)
```

**Q16 (2 marks).** Compare the top 3 most important variables to the three predictors you identified in Q1.
Do they match? Write one sentence about any variable that surprised you.
The top three most important variables were: lactate, MAP (mean arterial pressure) and WBC (white blood cell/leukocyte count). My top three predictors were lactate, heart rate and respiration rate. Lactate was the same, however mean arterial pressure makes sense as it relates to both heart rate and respiration. The thing that I did not account for however was leukocyte count. However, thinking about it and after looking up what leukocytes were, it makes sense that this would be an important factor. Leukocytes are the immune system's primary defense against infection, so knowing the count of them in the body can help determine if there is sepsis or not. If there are large amounts of leukocytes found in the body, then it is actively fighting off an infection, so this data would be quite useful when training models. 
---

## 8 Discussion

**Q17 (3 marks).** In 3–5 sentences, address both of the following:

- What is the single biggest limitation of this analysis (consider the label definition, the static sampling approach, generalisability to other hospitals, etc.)?
- Would you recommend deploying a model trained this way in a real ICU? If yes, under what monitoring conditions? If no, what would need to change first?

A1: The static sampling approach seems to be the biggest limitation. Sepsis is an illness that develops over time, so even if a patient is positive for sepsis, the time that we take the data from might show that their vitals are stable and they do not have sepsis. Having the continuous data, even if it would be a lot to sift through, would make the model more accurate and able to track changes over time rather than at random intervals. 
A2: I would not recommend this in an ICU setting. The main reason is because of the number of sepsis cases that it misses or misclassifies, which is more than the number of correctly classified sepsis cases. This would not be safe enough to use in such a life-or-death setting, it would need to be trained to be more sensitive. It also seems to not be great at certain demographics of people, where the PR-AUC of people ages 80+ was near 0, significantly different from the overall PR-AUC. This shows that the model might have some internal bias, potentially from the lack of data present. However, this would not do well in a hospital setting, espcially when the most in-risk groups are the ones that the models aren't great with.
---

## Session Info

```{r session-info}
sessionInfo()
```

## Credits

This practical is adapted from earlier versions of the course's medical databases lab and updated to use the [PhysioNet/CinC Challenge 2019 sepsis dataset](https://physionet.org/content/challenge-2019/1.0.0/) (Reyna et al., *Crit Care Med* 2020) to use a more modern `tidymodels` workflow.
